### 2017-01-25 real/draft.tex
[draft] p -> r
[draft] add comment on scale vs. rate par. for inv. gamma
[draft] comparison plot exact vs asymptotic
* remove old hedgehog plot
* remove paragraphs on conjugate model

\parindent=3ex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bm,amssymb,amsfonts,amsmath}  % bold math symbols, AMStex
\usepackage{graphicx}    % standard graphics
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\lleq}[1]{\label{#1} }
% TO REMOVE THE LABEL LETTERS FROM THE EQUATION DISPLAY, COMMENT OUT THIS LINE:
\renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }

@@

A major numerical simplification arises because of the $\Lumtot \to X$
transformation. Rewriting the \refeq{sll} formally as
\begin{align}
  \lleq{gmhb}
  p(Q  \cond n,\bml)
  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single-packet luminosity is narrow
in the sense that the standard deviation relative to the mean is at
the percent level. This implies that $p(Q \cond N, n, \bml)$ has one
mode from each $N$ and all modes are clearly separated. The
transformation $\Lumtot \to X$ effects a strong overlap between the
distributions $p(X \cond N, n, \bmx)$ with similar $N$, so
$p(X \cond n, \bmx)$ can be well approximated by a unimodal
distribution; see \reffig{hedgehog}.

\begin{figure}[ht]
  \centering
  % \includegraphics[width=0.48\textwidth]{hedgehog}
  \caption{Left: Prediction $p(\Lumtot \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
  \fred{I still need to pretty this up. I used the old binned approach
    with 50 samples in a bin from which I computed the mean and
    variance to plug into \refeq{gmr} but I computed the sum over $N$
    instead of the integral.}
\label{fig:hedgehog}
\end{figure}

@@

It remains to find the integrals over $\lambda$ and $\bmphi$ in
\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
distribution of \refeq{pms}. %
%
To determine $p(\bmphi\cond n,\bml)$, we note that the result
(\ref{rab}) also simplifies the calculation of the parameter posterior
$p(\bmphi\cond n,\bml)$ since, independently of the actual functional
model for $p(L\cond\bmphi)$ the only relevant information is contained
in the first moment $\mu$ and variance $\sigma^2$.  As set out in
Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
$\sigma^2$ implies that we can assign a Gaussian distribution to
$p(L\cond\bmphi)$
\begin{align}
  \lleq{rae}
  p(\Lum \cond \bmphi) =
  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
\end{align}
even though the actual data histograms may not be Gaussian. The reason
is that a Gaussian represents our state of knowledge \emph{before}
data $\bmy$ is observed. With this assignment,
$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
Theorem,
\begin{align}
  \lleq{rac}
  p(\mu,\sigma^2\cond n,\bml)
  &= \frac{
    \GaussianDist(\bml\cond\mu,\sigma^2)\;
    p(\mu\cond n)\;
    p(\sigma^2\cond n)}
  {p(\bml\cond n)}
\end{align}
\hans{I have inserted a uniform prior for $\mu$ here; see comments
  below. The original Gaussian prior has been retained below but I
  would suggest that the compliations introduced by it are
  unnecessary.}  %
In keeping with prior information that luminosities are always
positive, we assign an improper uniform prior
$p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
always, the posterior is independent of
$\mu_{\rm max}$.
\fred{only true if $\mu_{\rm max}$ large enough to contain posterior}
For $\sigma^2$, we use an $n$-independent Inverse Gamma
distribution prior with hyperparameters $(a_0,b_0)$ which reads
\begin{align}
  \lleq{rad}
  p(\sigma^2\cond a_0,b_0)
  = \InvGammaDist(\sigma^2\cond a_0,b_0)
  &= \frac{b_0^{a_0}}{\Gamma(a_0)} (\sigma^2)^{-a_0-1} \exp(-b_0/\sigma^2)
\end{align}
and is ``conjugate'' to the normal distribution likelihood of
\refeq{rae} in that the posterior for $(\mu,\sigma^2)$ is also a
Gaussian in $\mu$ and an Inverse Gamma in $\sigma^2$, but with
parameters updated by the data,
\begin{align}
  \lleq{rbd}
  p(\mu,\sigma^2\cond n,\bml,a_0,b_0)
  &= \GaussianDist(\mu\cond \mu_n,\sigma_n^2)
  \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \\
  \mu_n &= \langle\ell\rangle\\
  \sigma_n^2 &= \sigma^2/n \\
  a_n &= a_0 + \tfrac{1}{2}n \\
  b_n &= b_0 + \tfrac{1}{2} n\left( \langle\ell^2\rangle - \langle\ell\rangle^2\right)
\end{align}
with data sample moments $\langle\lum^k\rangle$ defined as $(1/n)
\sum_{j=1}^n \lum_j^k$.
%
Inserting Eqs.~(\ref{pms}) and (\ref{rbd}) into \refeq{asc}, we obtain
\begin{align}
  \lleq{ral}
  p(\Lumtot \cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
  \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
  \times\\
  &\quad\times
  \GammaDist(\lambda \cond n{-}a{+}1 , 1) \cdot
  \GaussianDist(\mu\cond \mu_n,\sigma_n^2) \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \;
  \nonumber
\end{align}
While the integral cannot be carried out analytically, the integrand
is unimodal and strongly peaked, so that the Laplace or saddle-point
approximation~\cite[Ch. 27]{mackay2003information} can be applied.
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hans{Some comments on the numerical values of hyperparameters $a$ in
  Poisson, $a,b$ in inverse gamma needed.} %
\hans{I would omit the text and calculations below. Apart from the
  extra parameters, the main reason is that the Gaussian prior in
  $\mu$ is not consistent with the fact that all luminosities are
  positive, while the gaussian permits negative $\mu$. This may not be
  important numerically, but the uniform prior in $\mu$ does
  circumvent this problem and is moreover less informative than the
  Gaussian prior. } %
\hans{Note also that letting $\mu_0{=}0$ and $ n_0\to 0$ below implies
  a prior gaussian of infinite width centered on zero, i.e.\ half of
  the values of $\mu$ will be negative. Could this be the reason why
  the numerics pick up a factor two?}%
\fred{No, what I used before was equivalent to a uniform prior that
  extends into the range $\mu<0$. The normalization cancels out and
  the negative region doesn't contribute. With the uniform prior, you
  assume that your upper cut-off on $\mu$ is large enough to contain
  the posterior but at the same time large values don't actually
  contribute so it doesn't matter if you include them or not. That
  means I can't pick up a factor of two there, it would only change
  the normalization but not the \emph{shape} of $p(Q)$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

To simplify the calculations, we assume a conjugate prior for unknown
$\mu$ and $\sigma^2$. The standard result for the posterior is a
Gaussian for $\mu$ and an Inverse Gamma distribution for $\sigma^2$,
\begin{align}
  \lleq{raf}
  p(\mu, \sigma^2 \cond \bmy)
  &= \GaussianDist\left(\mu \cond \mu_0^\prime, \sigma^2/n_0^\prime\right) %
    \InvGammaDist\left( \sigma^2 \,\Bigl|\, \tfrac{1}{2}\nu_0^\prime,
    \tfrac{1}{2}\nu_0^\prime \sigma_0^{\prime 2} \right)
\end{align}
where the hyperparameters $(n_0',\mu_0',\nu_0')$ are updated from
prior to posterior based on $\bmy$ with the sample mean
$\langle\lum\rangle = (1/n) \sum_{j=1}^n \lum_j$,
\begin{align}
  %\lleq{rag}
  \lleq{rah}  n_0^\prime
  &= n_0 + n,\\
  \lleq{rai}    \mu_0^\prime
  &= \frac{n_0 \mu_0 + n \langle\lum\rangle}{n_0^\prime},\\
  \lleq{raj}    \nu_0^\prime
  &= \nu_0 + n,\\
  \lleq{rak}
  \nu_0^\prime \sigma_0^{\prime 2}
  &= \nu_0 \sigma_0^{ 2}
    + \sum_{j=1}^n \left(\lum_j - \langle\lum\rangle\right)^2
    + \frac{n_0 n}{n_0^\prime} \left(\mu_0 - \langle\lum\rangle\right)^2\,.
\end{align}
By default, we take a noninformative prior and set $n_0 = \mu_0 =
\nu_0 = 0$, so the value $\sigma_0^2>0$ is irrelevant. %
  % The posterior
  % $ p(\expect{\Lum}, \expect{\Lum^2} \cond \bmy)$ is given by
  % \refeq{raf} with $\mu, \sigma^2$ replaced by
  % $\expect{\Lum}, \expect{\Lum^2}$ as indicated in \refeq{rae} because
  % the Jacobian determinant is one.
The above equations also show that the data $\bmy$ appear only in the
form of $n$ and the first two sample moments
$m_1 = \langle\ell\rangle$ and $m_2 = \langle\ell^2\rangle$.
Inserting all these ingredients into Eq.~(\ref{asc}), we obtain
\begin{align}
  \lleq{ral}
  p(\Lumtot \cond \bmy)
  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
    p(\lambda \cond n)\; p( \mu, \sigma^2 \cond \bmy)\;
    \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
    \,,
\end{align}
where $ p(\lambda \cond n)$ is given by \refeq{pms} and
$ p( \mu, \sigma^2 \cond \bmy)$ by Eqs.~(\ref{raf})--(\ref{rak}).

>>>>

\parindent=3ex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{bm,amssymb,amsfonts,amsmath}  % bold math symbols, AMStex
\usepackage{graphicx}    % standard graphics
\graphicspath{
  {./figures/}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\lleq}[1]{\label{#1} }
% TO REMOVE THE LABEL LETTERS FROM THE EQUATION DISPLAY, COMMENT OUT THIS LINE:
\renewcommand{\lleq}[1]{\label{#1} {\scriptstyle {\rm (#1)}} \hspace*{2ex} }

@@

A major numerical simplification arises because of the $\Lumtot \to X$
transformation. Rewriting the \refeq{sll} formally as
\begin{align}
  \lleq{gmhb}
  p(Q  \cond n,\bml)
  & = \sum_{N} p(N \cond n)\; p(Q \cond N, n, \bml),
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single-packet luminosity is narrow
in the sense that the standard deviation relative to the mean is at
the percent level. This implies that $p(Q \cond N, n, \bml)$ has one
mode from each $N$ and all modes are clearly separated. The
transformation $\Lumtot \to X$ effects a strong overlap between the
distributions $p(X \cond N, n, \bmx)$ with similar $N$, so
$p(X \cond n, \bmx)$ can be well approximated by a unimodal
distribution.

@@

It remains to find the integrals over $\lambda$ and $\bmphi$ in
\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
distribution of \refeq{pms}. %
%
To determine $p(\bmphi\cond n,\bml)$, we note that the result
(\ref{rab}) also simplifies the calculation of the parameter posterior
$p(\bmphi\cond n,\bml)$ since, independently of the actual functional
model for $p(L\cond\bmphi)$ the only relevant information is contained
in the first moment $\mu$ and variance $\sigma^2$.  As set out in
Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
$\sigma^2$ implies that we can assign a Gaussian distribution to
$p(L\cond\bmphi)$
\begin{align}
  \lleq{rae}
  p(\Lum \cond \bmphi) =
  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
\end{align}
even though the actual data histograms may not be Gaussian. The reason
is that a Gaussian represents our state of knowledge \emph{before}
data $\bmy$ is observed. With this assignment,
$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
Theorem,
\begin{align}
  \lleq{rac}
  p(\mu,\sigma^2\cond n,\bml)
  &= \frac{
    \GaussianDist(\bml\cond\mu,\sigma^2)\;
    p(\mu\cond n)\;
    p(\sigma^2\cond n)}
  {p(\bml\cond n)}
\end{align}
\hans{I have inserted a uniform prior for $\mu$ here; see comments
  below. The original Gaussian prior has been retained below but I
  would suggest that the compliations introduced by it are
  unnecessary.}  %
In keeping with prior information that luminosities are always
positive, we assign an improper uniform prior
$p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
always, the posterior is independent of
$\mu_{\rm max}$.
\fred{only true if $\mu_{\rm max}$ large enough to contain posterior}
For $\sigma^2$, we use an $n$-independent Inverse Gamma
distribution prior with hyperparameters $(a_0,b_0)$ which reads
\fred{Inconsistentcy. Now $b_0$ is a scale parameter but before $\beta$ was a \emph{rate} parameter in definition of $\Gamma$}
\begin{align}
  \lleq{rad}
  p(\sigma^2\cond a_0,b_0)
  = \InvGammaDist(\sigma^2\cond a_0,b_0)
  &= \frac{b_0^{a_0}}{\Gamma(a_0)} (\sigma^2)^{-a_0-1} \exp(-b_0/\sigma^2)
\end{align}
and is ``conjugate'' to the normal distribution likelihood of
\refeq{rae} in that the posterior for $(\mu,\sigma^2)$ is also a
Gaussian in $\mu$ and an Inverse Gamma in $\sigma^2$, but with
parameters updated by the data,
\begin{align}
  \lleq{rbd}
  p(\mu,\sigma^2\cond n,\bml,a_0,b_0)
  &= \GaussianDist(\mu\cond \mu_n,\sigma_n^2)
  \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \\
  \mu_n &= \langle\ell\rangle\\
  \sigma_n^2 &= \sigma^2/n \\
  a_n &= a_0 + \tfrac{1}{2}n \\
  b_n &= b_0 + \tfrac{1}{2} n\left( \langle\ell^2\rangle - \langle\ell\rangle^2\right)
\end{align}
with data sample moments $\langle\lum^k\rangle$ defined as $(1/n)
\sum_{j=1}^n \lum_j^k$.
%
Inserting Eqs.~(\ref{pms}) and (\ref{rbd}) into \refeq{asc}, we obtain
\begin{align}
  \lleq{ral}
  p(\Lumtot \cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
  \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
  \times\\
  &\quad\times
  \GammaDist(\lambda \cond n{-}a{+}1 , 1) \cdot
  \GaussianDist(\mu\cond \mu_n,\sigma_n^2) \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \;
  \nonumber
\end{align}
While the integral cannot be carried out analytically, the integrand
is unimodal and strongly peaked, so that the Laplace or saddle-point
approximation~\cite[Ch. 27]{mackay2003information} can be applied.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.48\textwidth]{asymptotic}
  \caption{Numerical prediction of $\Lumtot$ with the sum over $N$
    from \refeq{spk} or the asymptotic expression from \refeq{ral} for $n$
    samples from $\GammaDist(\alpha=1.5, \beta=60)$}
\label{fig:asymptotic}
\end{figure}

\fred{In \reffig{asymptotic}, we contrast the asymptotic expression
  with the ``exact'' expression \refeq{spk} where the integrals over
  $\bmphi$ are done numerically to a requested relative accuracy of
  $10^{-5}$ through the Cubature
  package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
  computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values
  and checked if the probability integrates to one using Simpson's
  rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
  ($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
  $n=10$, the large discrepancy arises mostly because the constraint
  $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in \refeq{ral}
  whereas it is satisfied by
  \refeq{spk}. Following~\cite{tierney_accurate_1986}, we renormalized
  the results to exactly integrate to one for ease of comparison in
  \reffig{asymptotic}. Note that the lack of normalization may not be
  relevant; e.g. if \refeq{ral} is used as the likelihood for a single
  bin in Bayesian parameter inference.

  We verified that for $n>80$, the two methods agree even better than
  shown there and that the normalization of the asymptotic result
  approaches one. }

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hans{Some comments on the numerical values of hyperparameters $a$ in
  Poisson, $a,b$ in inverse gamma needed.} %
\hans{I would omit the text and calculations below. Apart from the
  extra parameters, the main reason is that the Gaussian prior in
  $\mu$ is not consistent with the fact that all luminosities are
  positive, while the gaussian permits negative $\mu$. This may not be
  important numerically, but the uniform prior in $\mu$ does
  circumvent this problem and is moreover less informative than the
  Gaussian prior. } %
\hans{Note also that letting $\mu_0{=}0$ and $ n_0\to 0$ below implies
  a prior gaussian of infinite width centered on zero, i.e.\ half of
  the values of $\mu$ will be negative. Could this be the reason why
  the numerics pick up a factor two?}%
\fred{No, what I used before was equivalent to a uniform prior that
  extends into the range $\mu<0$. The normalization cancels out and
  the negative region doesn't contribute. With the uniform prior, you
  assume that your upper cut-off on $\mu$ is large enough to contain
  the posterior but at the same time large values don't actually
  contribute so it doesn't matter if you include them or not. That
  means I can't pick up a factor of two there, it would only change
  the normalization but not the \emph{shape} of $p(Q)$}
\fred{The error is fixed and I remove the next sections as Hans suggested.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
